---
title: "Liver bud development"
author: "Yutong Wang"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

# Introduction

We analyze the liver bud dataset from [Camp et al 2017](https://www.nature.com/articles/nature22796).

Everything below is totally self-contained: you can copy-paste word for word and it will run. The single-line code block in section _CorGI unsupervised feature selection_ takes one hour to run. Everything else in total takes under one minute.

This vignette contains long sections headed by _Reproducing paper figures_. These sections are not necessary for using CorGI. To see a more user-friendly pipeline for applying CorGI to your research, see [Get Started](https://yutongwangumich.github.io/corgi/articles/corgi.html).

The section under _Exploratory analysis_ (and in particular, normalization by making the library sizes comparable across batches) is also not necessary to use CorGI. In fact, it makes no difference whether the counts are integers, log-transformed, TPM, FPKM, etc. 


```{r}
library(corgi) 
library(knitr) # for creating pretty tables
library(ggplot2) # for creating pretty plots
library(cowplot) # for creating pretty panel of plots
library(dplyr) # for creating pretty code
library(forcats) # for concatenating factors
library(scmap) # for the compared feature selection method
```

# Data cleaning

Download from GSE and view the top-left corner of the data table
```{r}
con <- gzcon(
  url("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE81nnn/GSE81252/suppl/GSE81252_data.cast.log2.liverbud.csv.gz")
  )
txt <- readLines(con)
liverbud <- read.csv(file = textConnection(txt))
kable(liverbud[1:10,1:6])
```

Note:

- The first three columns are metadata while the rest are the log2 expressions

- The column `assignment_LB` contains both the lineage and batch information

- EC = endothelial cells, HE = hepatic endoderm, MC = mesenchymal cells

- IN = input cell lines, LB = liver bud

```{r}
metadata <- liverbud[,1:3]
counts <- t(as.matrix(liverbud[,4:ncol(liverbud)]))
colnames(counts) <- metadata$cell_id
rownames(metadata) <- metadata$cell_id
metadata$LB_status <- 
  unlist(lapply(X = metadata$assignment_LB,
                FUN = function(x) ifelse(test = grepl("LB",x),yes = "Liver_bud",no = "Input")
                ))
unique(metadata$LB_status)

metadata$lineage <- 
  factor(unlist(
    lapply(X = metadata$assignment_LB,
           FUN = function(x) strsplit(as.character(x),split = "-",fixed = T)[[1]][1])
    ))
levels(metadata$lineage)

plyr::mapvalues(x = metadata$lineage,
                from = levels(metadata$lineage),
                to = c("Endothelial","Hepatic","Mesenchymal")) -> metadata$lineage
```


# Exploratory analysis

Number of cells in each batch in each lineage
```{r}
kable(table(metadata$lineage,metadata$LB_status))
```


Convert to `SingleCellExperiment` objects
```{r}
library(scran)
camp <- SingleCellExperiment(assays = list(logcounts = counts),colData = metadata)

IN <- camp[,camp$LB_status=="Input"]
LB <- camp[,camp$LB_status=="Liver_bud"]
```

Normalization by making the library sizes comparable across batches
```{r}
out <- scran::multiBatchNorm(IN,LB,assay.type = "logcounts")
IN <- out[[1]]
LB <- out[[2]]
```

Combine the datasets
```{r}
combined <- cbind(logcounts(IN),logcounts(LB))
lineage <- fct_c(IN$lineage,LB$lineage)
batch <- c(rep("Input",ncol(IN)),rep("Liver bud",ncol(LB)))
```


## PC plot using all the genes

```{r}
pc <- prcomp(t(combined),rank. = 2)$x
qplot(pc[,1],pc[,2],color = lineage,shape = batch)
```


# CorGI unsupervised feature selection

```{r, eval = F}
corgi_output_liver_bud <- run_corgi(logcounts(IN),logcounts(LB))
```

```{r, eval = F, include= F}
devtools::use_data(corgi_output_liver_bud,overwrite = T)
```

```{r,include = F}
data("corgi_output_liver_bud")
```

## Selecting the top CorGI genes

```{r}
corgi_gene_set <- select_top_corgi_genes(corgi_output_liver_bud,200)
```


## PCA on CorGI genes


```{r}
pc <- prcomp(t(combined[corgi_gene_set,]),rank. = 2)$x
qplot(pc[,1],pc[,2],color = lineage,shape = batch)
```


# Reproducing paper figures

Gene sets to compare CorGI against

```{r}
HDG_ranking <- function(sce){
  rowData(sce)$feature_symbol <- rownames(sce)
  sce <- selectFeatures(sce)
  return(rownames(sce)[order(rowData(sce)[["scmap_scores"]],decreasing = T,na.last = T)])
}

gene_sets <- get_compared_gene_sets(
  batch1_top_genes = HDG_ranking(IN),
  batch1_name = "HDG(Input)",
  batch2_top_genes = HDG_ranking(LB),
  batch2_name = "HDG(Liver bud)",
  desired_size = length(corgi_gene_set)
)

gene_sets[["CorGI"]] <- corgi_gene_set

gene_sets <- gene_sets[6-(1:5)]
lapply(gene_sets, length)
```


```{r}
embeddings_pca <- 
  lapply(
    X = gene_sets,
    FUN = function(gene_set) {
      prcomp(t(combined[gene_set,]),rank. = 2)$x
    }
  )
```


```{r}
my_shape_palette <- scale_shape_manual(values = c(1, 16))
my_color_palette <- scale_color_manual(values = c("#E69F00","#000000", "#56B4E9"))
```


```{r}
get_scatterplots <- function(embeddings, batch, cell_type){
  lapply(X = names(gene_sets),
         FUN = function(gs_name) {
           emb <- embeddings[[gs_name]]
           plot_dimensionality_reduction(emb, batch, cell_type) +
             my_shape_palette +
             my_color_palette + 
             ggtitle(paste0(gs_name, ", ", round(corgi::batch_mixing(emb, batch), 2)))
         })
}

get_AUC <- function(embeddings, cell_type, cell_type_pred, train, test){
  knn_results <-
    Reduce(rbind,lapply(X = names(embeddings),
                        FUN = function(gs_name){
                          emb <- embeddings[[gs_name]]
                          results <- corgi::cluster_coherence(emb, cell_type, cell_type_pred, train, test)
                          results$Gene_set <- gs_name
                          return(results)
                        }))
  return(knn_results)
}
```

```{r}
emb_name <- "PC"
embeddings <- embeddings_pca
get_scatterplots(embeddings,
                 batch,
                 lineage) -> dr_plots

plot_grid(
  plot_grid(
    get_axes_legend(emb_name),
    get_shape_legend(batch,my_shape_palette)
  ),
  get_color_legend(lineage,my_color_palette),
  nrow = 2) -> legends

dr_plots[[length(dr_plots)+1]] <- legends

dr_plots[["nrow"]] <- 2

dr_plots[["labels"]] <- c("A","B","C","D","E","")

do.call(plot_grid,dr_plots) -> top_panel
top_panel
```

```{r}
train <- batch == "Input"
test <- batch == "Liver bud"

# AUC plots
lapply(
  X = c("Endothelial","Hepatic","Mesenchymal"),
  FUN =
    function(ct) {
      df <- get_AUC(embeddings, lineage, ct, train, test)
      df$Gene_set <- factor(df$Gene_set,levels = names(embeddings))
      ggplot(df, 
             aes(x = FPR, y = TPR)) + 
        geom_point(aes(color = Gene_set, shape = Gene_set),size = 3) +
        scale_shape_manual(values = 65:70) + 
        theme_bw() +
        theme(legend.position = "bottom") +
        scale_x_continuous(limits = c(0,1)) +
        scale_y_continuous(limits = c(0,1)) +
        ggtitle(ct)
    }
) -> AUC_plts

AUC_legends <- get_legend(AUC_plts[[1]])

lapply(X = AUC_plts, FUN = function(p) p + theme(legend.position = "none")) -> AUC_plts

AUC_plts[["nrow"]] <- 1
AUC_plts[["labels"]] <- c("F","G","H")

plot_grid(
  do.call(what = plot_grid, args = AUC_plts),
  AUC_legends,
  nrow = 2,
  rel_heights = c(10,1)) -> bottom_panel

bottom_panel
```

```{r,include=F,eval=F}
liver_bud_panel <- plot_grid(top_panel,bottom_panel,nrow=2,rel_heights = c(2,1))

ggsave(filename = "liver_bud.eps",device = "eps",plot = liver_bud_panel, path = "~/CorGI_figures",width = 8,height = 8)
```





# Out of sample alignment of human liver cells

In the previous sections, we ran CorGI feature selection on the input and liver bud cells. We consider the human liver cells sequenced also by Camp et al. We analyze how genes obtained from the previous section can be used to gain insight in the new data.

## Data cleaning
```{r}
con <- gzcon(
  url("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE96nnn/GSE96981/suppl/GSE96981_data.human.liver.csv.gz")
  )
txt <- readLines(con)
human_liver <- read.csv(file = textConnection(txt))

```

The first 13 columns are metadata
```{r}
colnames(human_liver)[1:13]
mdat <- human_liver[,1:13]
```

The remaining are genes
```{r}
colnames(human_liver)[14:20]
counts <- t(as.matrix(human_liver[,14:ncol(human_liver)]))
colnames(counts) <- mdat$cell_id
```

Let us refer back to [Camp et al. 2017, Figure 3b](https://www.nature.com/articles/nature22796#f3) for the cluster labels:
```{r}
mdat$ident <- as.factor(mdat$ident)
qplot(mdat$tSNE_1,mdat$tSNE_2,color = mdat$ident)
```

Just to double check...
```{r}
mdat$cell_type_raw <- 
  plyr::mapvalues(mdat$ident,
                  from = 1:7,
                  to = c("Lymphoblast",
                         "Erythroblast",
                         "Adult hepatocyte",
                         "Mesenchymal",
                         "Endothelial",
                         "Fetal hepatocyte",
                         "Kupffer"))
qplot(mdat$tSNE_1,mdat$tSNE_2,color = mdat$cell_type_raw)
```

Collapse some of the cell types for visualization
```{r}
mdat$lineage_show <- 
  plyr::mapvalues(mdat$ident,
                  from = 1:7,
                  to = c("Other",
                         "Other",
                         "Adult hepatocyte",
                         "Mesenchymal",
                         "Endothelial",
                         "Fetal hepatocyte",
                         "Other"))
```


Collapse _Adult hepatocyte_ and _Immature hepatocyte_ to simply _Hepatic_ for the KNN classification part
```{r}
mdat$lineage_predict <- 
  plyr::mapvalues(mdat$ident,
                  from = 1:7,
                  to = c("Other",
                         "Other",
                         "Hepatic",
                         "Mesenchymal",
                         "Endothelial",
                         "Hepatic",
                         "Other"))
```



```{r}
HL <- SingleCellExperiment(assays = list(logcounts = counts),colData = mdat)
```


# Reproducing paper figures

Combine the datasets and normalize
```{r}
shared_genes <- intersect(rownames(IN),rownames(HL))
IN <- IN[shared_genes,]
LB <- LB[shared_genes,]
HL <- HL[shared_genes,]
out <- scran::multiBatchNorm(IN,LB,HL,assay.type = "logcounts")
IN <- out[[1]]
LB <- out[[2]]
HL <- out[[3]]

lineage_show <- fct_c(IN$lineage,LB$lineage,HL$lineage_show)
lineage_pred <- fct_c(IN$lineage,LB$lineage,HL$lineage_predict)
batch <- c(rep("Input",ncol(IN)),rep("Liver bud",ncol(LB)),rep("Primary",ncol(HL)))
```


```{r}
my_shape_palette <- scale_shape_manual(values = c(1, 16,2))
my_color_palette <- scale_color_manual(values = c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2"))
```


```{r}
embeddings_mnn <- 
  lapply(
    X = gene_sets,
    FUN = function(gene_set) {
      mnn.out <- mnnCorrect(logcounts(IN[gene_set,]),
                            logcounts(LB[gene_set,]),
                            logcounts(HL[gene_set,]))
      t.mnn <- as.matrix(t(do.call(cbind, mnn.out$corrected)))
      pca.mnn <- prcomp(t.mnn, rank=2)
      return(pca.mnn$x)
    }
  )
```



```{r}
unique(lineage_show)
plyr::mapvalues(
  x = lineage_show,
  from = unique(lineage_show),
  to = c("Hep.","End.","Mes.", "Fetal hep.", "Other", "Adult hep.")
) -> lineage_show
```


```{r}
emb_name <- "MNNPC"
embeddings <- embeddings_mnn
get_scatterplots(embeddings,
                 batch,
                 lineage_show) -> dr_plots

plot_grid(
  plot_grid(
    get_axes_legend(emb_name),
    get_shape_legend(batch,my_shape_palette)
  ),
  get_color_legend(lineage_show,my_color_palette,ncol = 2),
  nrow = 2) -> legends

dr_plots[[length(dr_plots)+1]] <- legends

dr_plots[["nrow"]] <- 2

dr_plots[["labels"]] <- c("A","B","C","D","E","")

do.call(plot_grid,dr_plots) -> top_panel
top_panel
```

```{r}
train <- batch %in% c("Input","Liver bud")
test <- batch == "Primary"

# AUC plots
lapply(
  X = c("Endothelial","Hepatic","Mesenchymal"),
  FUN =
    function(ct) {
      df <- get_AUC(embeddings, lineage_pred, ct, train, test)
      df$Gene_set <- factor(df$Gene_set,levels = names(embeddings))
      ggplot(df, 
             aes(x = FPR, y = TPR)) + 
        geom_point(aes(color = Gene_set, shape = Gene_set),size = 3) +
        scale_shape_manual(values = 65:70) + 
        theme_bw() +
        theme(legend.position = "bottom") +
        scale_x_continuous(limits = c(0,1)) +
        scale_y_continuous(limits = c(0,1)) +
        ggtitle(ct)
    }
) -> AUC_plts

AUC_legends <- get_legend(AUC_plts[[1]])

lapply(X = AUC_plts, FUN = function(p) p + theme(legend.position = "none")) -> AUC_plts

AUC_plts[["nrow"]] <- 1
AUC_plts[["labels"]] <- c("F","G","H")

plot_grid(
  do.call(what = plot_grid, args = AUC_plts),
  AUC_legends,
  nrow = 2,
  rel_heights = c(10,1)) -> bottom_panel

bottom_panel
```

```{r,include=F,eval=F}
human_liver_panel <- plot_grid(top_panel,bottom_panel,nrow=2,rel_heights = c(2,1))

ggsave(filename = "human_liver.eps",device = "eps",plot = human_liver_panel, path = "~/CorGI_figures",width = 8,height = 8)
```

