---
title: "Neurogenesis in the SVZ"
author: "Yutong Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

In this vignette, we the semi-supervised version of `corgi` applied to the neurogenesis in SVZ datasets:

- [Dulken et al. 2017](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5269583/)
- [Llorens-Bobadilla et al. 2015](https://www.ncbi.nlm.nih.gov/pubmed/26235341)

```{r}
library(corgi)
library(knitr)
library(ggplot2)
```

# Loading Llorens-Bobadilla et al. 2015 dataset

Load data from Llorens-Bobadilla. This code is obtained from the [GitHub code for Figure 6B in Dulken et al 2017](https://github.com/bdulken/SVZ_NSC_Dulken_2/blob/master/Figure6_FigureS6/Llorens%20Projections%20_%20Figure%206B%2C%20S6A%2C%20S6B.R)

```{r}
Llorens_allcounts<-read.table(url("https://github.com/bdulken/SVZ_NSC_Dulken_2/raw/master/Files/Llorens_counts_allgenes.txt"))

allcounts_allcells<-Llorens_allcounts

#Remove neuroblasts from Llorenss data
#allcounts_allcells_notaps<-allcounts_allcells[!grepl("tap",colnames(allcounts_allcells))]
allcounts_allcells_noblasts<-allcounts_allcells[!grepl("PSA",colnames(allcounts_allcells))]

#Remove oligodendrocytes from Llorenss data
oligos<-as.vector(read.table(url("https://raw.githubusercontent.com/bdulken/SVZ_NSC_Dulken_2/master/Files/comp_oligos.txt"))[,1])
allcounts_allcells_noblasts_nooligo<-allcounts_allcells_noblasts[,-na.omit(match(oligos,colnames(allcounts_allcells_noblasts)))]
allcounts_allcells_noblasts_nooligo_noERCC<-allcounts_allcells_noblasts_nooligo[!grepl("ERCC-",rownames(allcounts_allcells_noblasts_nooligo)),]

llorens <- allcounts_allcells_noblasts_nooligo_noERCC

llorens_cell_type <-  factor(
  unlist(
    lapply(X = colnames(llorens),
           FUN = function(x){substr(x,1,1)})
    )
  )

# Make the cluster labels more intuitive
# neural stem and transient transit amplifying cells (NSC and TAP)
llorens_cell_type <- plyr::mapvalues(x = llorens_cell_type,
                                     from = c("N","t"),
                                     to = c("NSC","TAP")) 
```

# Loading Dulken et al 2017 dataset

This code is obtained from the [GitHub code for Figure 1 in Dulken et al 2017](https://github.com/bdulken/SVZ_NSC_Dulken_2/blob/master/Figure1/PCA%20-%20Outliers%20Removed%20_%20Figure%201C.R). The difference between our code and theirs is that we don't filter out cell-cycle genes.

```{r}
#Loading all high quality cells and filtering for lowly expressed genes
spec_pops<-read.table(url("https://raw.githubusercontent.com/bdulken/SVZ_NSC_Dulken_2/master/Files/AllCounts_specPops_read_gene_ERCC_filt_FINAL.txt"))

allcounts_allcells<-spec_pops

#Removing Oligodendrocytes and Outliers
oligos<-as.vector(read.table(url("https://raw.githubusercontent.com/bdulken/SVZ_NSC_Dulken_2/master/Files/STAR_oligos_updated_09232015.txt"))[,1])
allcounts_allcells_nooligo<-allcounts_allcells[,-na.omit(match(oligos,colnames(allcounts_allcells)))]

#Filtering for expressed by 5 cells at 10 counts
greaterthan0<-allcounts_allcells_nooligo>10
greaterthan0sum<-rowSums(greaterthan0)
allcounts_allcells_nooligo_genefilt<-allcounts_allcells_nooligo[greaterthan0sum>=5,]
dulken <- allcounts_allcells_nooligo_genefilt

dulken_cell_type <- 
  factor(
    unlist(
      lapply(X=colnames(dulken),
             FUN=function(x){strsplit(x,split = "_",fixed=T)[[1]][1]})
      )
    )
  
```



We consider only genes in both datasets.
```{r}
shared_genes <- intersect(rownames(dulken),rownames(llorens))
length(shared_genes)
```


Combine the datasets

```{r}
dulken <- dulken[shared_genes,]
llorens <- llorens[shared_genes,]
comb <- cbind(dulken,llorens)
```

Metadata for the merged dataset
```{r}
cell_type <- forcats::fct_c(dulken_cell_type,llorens_cell_type)
batch <- c(rep("Dulken",ncol(dulken)),rep("Llorens-Bobadilla",ncol(llorens)))

kable(table(cell_type,batch))
```


Download the genes ranked by their supervised feature selection method
```{r}
library("openxlsx")
consensus_genes <- read.xlsx("https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5269583/bin/NIHMS839312-supplement-7.xlsx")
dim(consensus_genes)
kable(consensus_genes[1:10,1:5]) # print a nice looking table of the top-left corner of the data

```

```{r}
consensus_genes <- consensus_genes[,-c(1)] # remove the first column, which is metadata

# Input: percentage (a number between 0 and 100)
# Output: gene_set consisting of genes showing up in at least percentage number of models
top_consensus_genes <- function(percentage) {
  gene_set <- c()
  
  for (g in Reduce(union, consensus_genes)) {
    if (sum(sapply(
      X = consensus_genes,
      FUN = function(model) {
        g %in% model
      })) > percentage) {
      gene_set <- c(gene_set, g)
    }
  }
  return(gene_set)
}

consensus34 <- top_consensus_genes(percentage = 50)

```

There are 34 genes that show up in more than 50% of the models, in agreement with Dulken et al. 2017

```{r}
length(consensus34)
```

# Running semi-supervised corgi with the consensus gene set as markers


```{r,eval = F}
corgi_output_neurogenesis <- run_corgi(dulken,llorens,time_per_phase = 10*60,must_have_genes = consensus34)
```

```{r,eval=F,include=F}
devtools::use_data(corgi_output_neurogenesis)
```
```{r,include=F}
data("corgi_output_neurogenesis")
```


```{r}
corgi_gene_set <- select_top_corgi_genes(corgi_output_neurogenesis,n = 100)
length(corgi_gene_set)


D <- (1-cor(comb[union(corgi_gene_set,consensus34),],method = "spearman"))/2
mds <- cmdscale(D,k=2)

qplot(mds[,1],mds[,2],color = cell_type,shape = batch) + 
  scale_shape_manual(values=c(1,16))
```


Combine some of the cell-types for visualization
```{r}
cell_type2 <- plyr::mapvalues(x = cell_type,
                              from = c("aNSC", "Ast", "NPC", "qNSC", "NSC", "TAP"),
                              to = c("NSC","Ast","NPC","NSC","NSC","NPC"))

qplot(mds[,1],mds[,2],color = cell_type2,shape = batch) + 
  scale_shape_manual(values=c(1,16))

```
