---
title: "Neurogenesis in the SVZ"
author: "Yutong Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

# Introduction

In this vignette, we use the _semi-supervised_ version of `corgi` applied to the neurogenesis in subventricular zone datasets:

- [Dulken et al. 2017](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5269583/)
- [Llorens-Bobadilla et al. 2015](https://www.ncbi.nlm.nih.gov/pubmed/26235341)

```{r}
library(corgi)
library(knitr)
library(ggplot2)
library(scran) # we only need to use the multiBatchNorm function
library(scmap) # we only need to use the selectFeatures function
library(cowplot) # we use this to generate the multipanel plots
library(dplyr)

```

# Data cleaning

Load data from Llorens-Bobadilla et al. This code is obtained from the [GitHub code for Figure 6B in Dulken et al 2017](https://github.com/bdulken/SVZ_NSC_Dulken_2/blob/master/Figure6_FigureS6/Llorens%20Projections%20_%20Figure%206B%2C%20S6A%2C%20S6B.R)

```{r}
Llorens_allcounts<-read.table(url("https://github.com/bdulken/SVZ_NSC_Dulken_2/raw/master/Files/Llorens_counts_allgenes.txt"))

allcounts_allcells<-Llorens_allcounts

#Remove neuroblasts from Llorenss data
#allcounts_allcells_notaps<-allcounts_allcells[!grepl("tap",colnames(allcounts_allcells))]
allcounts_allcells_noblasts<-allcounts_allcells[!grepl("PSA",colnames(allcounts_allcells))]

#Remove oligodendrocytes from Llorenss data
oligos<-as.vector(read.table(url("https://raw.githubusercontent.com/bdulken/SVZ_NSC_Dulken_2/master/Files/comp_oligos.txt"))[,1])
allcounts_allcells_noblasts_nooligo<-allcounts_allcells_noblasts[,-na.omit(match(oligos,colnames(allcounts_allcells_noblasts)))]
allcounts_allcells_noblasts_nooligo_noERCC<-allcounts_allcells_noblasts_nooligo[!grepl("ERCC-",rownames(allcounts_allcells_noblasts_nooligo)),]

llorens <- allcounts_allcells_noblasts_nooligo_noERCC
```

Overview of the Llorens-Bobadilla dataset
```{r}
kable(llorens[1:10,1:6])
```

Parse the cell type metadata
```{r}
llorens_cell_type <-  factor(
  unlist(
    lapply(X = colnames(llorens),
           FUN = function(x){substr(x,1,1)})
    )
  )
```


Make the cluster labels more intuitive
neural stem and transient transit amplifying cells (NSC and TAP)

```{r}
llorens_cell_type <- plyr::mapvalues(x = llorens_cell_type,
                                     from = c("N","t"),
                                     to = c("NSC","TAP")) 
```


Load Dulken et al 2017 dataset

This code is obtained from the [GitHub code for Figure 1 in Dulken et al 2017](https://github.com/bdulken/SVZ_NSC_Dulken_2/blob/master/Figure1/PCA%20-%20Outliers%20Removed%20_%20Figure%201C.R). The difference between our code and theirs is that we don't filter out cell-cycle genes.

```{r}
#Loading all high quality cells and filtering for lowly expressed genes
spec_pops<-read.table(url("https://raw.githubusercontent.com/bdulken/SVZ_NSC_Dulken_2/master/Files/AllCounts_specPops_read_gene_ERCC_filt_FINAL.txt"))

allcounts_allcells<-spec_pops

#Removing Oligodendrocytes and Outliers
oligos<-as.vector(read.table(url("https://raw.githubusercontent.com/bdulken/SVZ_NSC_Dulken_2/master/Files/STAR_oligos_updated_09232015.txt"))[,1])
allcounts_allcells_nooligo<-allcounts_allcells[,-na.omit(match(oligos,colnames(allcounts_allcells)))]

#Filtering for expressed by 5 cells at 10 counts
greaterthan0<-allcounts_allcells_nooligo>10
greaterthan0sum<-rowSums(greaterthan0)
allcounts_allcells_nooligo_genefilt<-allcounts_allcells_nooligo[greaterthan0sum>=5,]
dulken <- allcounts_allcells_nooligo_genefilt
```

Overview of the Dulken dataset
```{r}
kable(dulken[1:10,1:6])
```

Parse the cell type metadata

```{r}
dulken_cell_type <- 
  factor(
    unlist(
      lapply(X=colnames(dulken),
             FUN=function(x){strsplit(x,split = "_",fixed=T)[[1]][1]})
      )
    )
  
```



We consider only genes in both datasets.
```{r}
shared_genes <- intersect(rownames(dulken),rownames(llorens))
length(shared_genes)
dulken <- dulken[shared_genes,]
llorens <- llorens[shared_genes,]
```

# Exploratory analysis

Combine the datasets
```{r}
comb <- cbind(dulken,llorens)
```


Merged metadata for the merged dataset
```{r}
cell_type <- forcats::fct_c(dulken_cell_type,llorens_cell_type)
batch <- c(rep("Dulken",ncol(dulken)),rep("Llorens",ncol(llorens)))
```

Overview of the cell types in each dataset
```{r}
kable(table(cell_type, batch))
```

Simplify and combine some of the cell-types for visualization
```{r}
cell_type2 <- plyr::mapvalues(x = cell_type,
                              from = c("aNSC", "Ast", "NPC", "qNSC", "NSC", "TAP"),
                              to = c("NSC","Ast","NPC","NSC","NSC","NPC"))
kable(table(cell_type2, batch))
```

```{r}
my_shape_palette <- c(1, 16)
my_color_palette <- c("#E69F00", "#000000", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")


qplot <- function(...){
  ggplot2::qplot(...) +
    scale_shape_manual(values = my_shape_palette) +
    scale_color_manual(values = my_color_palette)
}
```


## Dimensionality reduction using all genes

Multidimensional-scaling on Spearman rho distance matrix using all the genes results in a scatterplot with clearly separated batches

```{r}
D <- sqrt((1-cor(comb,method = "spearman"))/2) # Spearman rho distance matrix
mds <- cmdscale(D,k=2) # Multidimensional-scaling

qplot(mds[,1],mds[,2],color = cell_type,shape = batch)

qplot(mds[,1],mds[,2],color = cell_type2,shape = batch) 
```


Download the genes ranked by their supervised feature selection method
```{r}
library("openxlsx")
consensus_genes <- read.xlsx("https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5269583/bin/NIHMS839312-supplement-7.xlsx")
dim(consensus_genes)
kable(consensus_genes[1:10,1:5]) # print a nice looking table of the top-left corner of the data
```

Note:

- The table `consensus_genes` has 101 columns, the first column is metadata
- The remainder columns are gene rankings computed by Dulken et al.
- Each column is the top 100 genes produced by one computed model.
- A gene that shows up in multiple columns means that there is a _consensus_ across multiple models that the gene should be ranked in the top 100, hence the name.

```{r}
consensus_genes <- consensus_genes[,-c(1)] # remove the first column, which is metadata
```

```{r}

all_consensus_genes <- intersect(Reduce(union, consensus_genes),
                                 shared_genes)

all_consensus_genes %>%
  sapply(function(g){
    # in how many models does g show up in?
    consensus_genes %>%
      sapply(function(model) {
        g %in% model
      }) %>%
      sum
  }) %>% 
  sort(decreasing = T) ->
  top_consensus_genes
```

The vector `top_consensus_genes` counts how many times a given gene show up in the models.
```{r}
head(top_consensus_genes)
```

```{r}
consensus34 <- names(which(top_consensus_genes>50))
```

There are 34 genes that show up in more than 50% of the models. This is the _consensus-ordering_ gene set used in Dulken et al. 2017.

```{r}
length(consensus34)
```

# Semi-supervised CorGI with the consensus genes as markers

```{r,eval = F}
set.seed(0)
corgi_output_neurogenesis <- run_corgi(dulken, llorens, must_have_genes = consensus34)
```

```{r,eval = F,include=F}
devtools::use_data(corgi_output_neurogenesis,overwrite = T)
```

```{r,include=F}
data("corgi_output_neurogenesis")
```

## Dimensionality reduction using MDS
```{r}
corgi_gene_set <- select_top_corgi_genes(corgi_output_neurogenesis,n = 100, phase_use = 1)
length(corgi_gene_set)

D <- (1-cor(comb[union(corgi_gene_set,consensus34),],method = "spearman"))/2
mds <- cmdscale(D,k=2)

qplot(mds[,1],mds[,2],color = cell_type,shape = batch)
```


# Recreating the paper figure

In this section, we recreate the figures in our paper, i.e., comparing with highly dropped-out genes.

## Normalize the datasets
```{r}
dulken <- SingleCellExperiment(assay = list(counts = as.matrix(dulken)))
llorens <- SingleCellExperiment(assay = list(counts = as.matrix(llorens)))

out <- multiBatchNorm(dulken,llorens) # the output has logcounts 

dulken <- out[[1]]
llorens <- out[[2]]
```



```{r}
HDG_ranking <- function(sce){
  rowData(sce)$feature_symbol <- rownames(sce)
  sce <- selectFeatures(sce)
  return(rownames(sce)[order(rowData(sce)[["scmap_scores"]],decreasing = T,na.last = T)])
}

n <- length(union(corgi_gene_set,consensus34))

gene_sets <- get_compared_gene_sets(
  batch1_top_genes = HDG_ranking(dulken),
  batch1_name = "HDG(Dulken)",
  batch2_top_genes = HDG_ranking(llorens), 
  batch2_name = "HDG(Llorens)",
  desired_size = n,
  marker_genes = consensus34)

gene_sets[["C134"]] <- names(top_consensus_genes)[1:n]

gene_sets[["CorGI"]] <- union(corgi_gene_set,consensus34)

gene_sets <- gene_sets[c(6,5,4,3,2,1)]
lapply(gene_sets, length)
```

scmapCluster parameters

```{r}
rowData(dulken)$feature_symbol <- rownames(dulken)
rowData(llorens)$feature_symbol <- rownames(llorens)
dulken$cell_type1 <- cell_type2[batch == "Dulken"]
llorens$cell_type1 <- cell_type2[batch == "Llorens"]

query <- llorens
reference <- dulken[,dulken$cell_type1!="Ast"]
threshold <- 0.3
```


```{r}
gene_set <- gene_sets[["CorGI"]]
corgi::run_scmap(
  query = query,
  ref = reference,
  gene_set = gene_set,
  threshold = threshold
)
```


```{r}
gene_set <- gene_sets[["C134"]]
corgi::run_scmap(
  query = query,
  ref = reference,
  gene_set = gene_set,
  threshold = threshold
)
```



```{r}
comparison_results <-
  run_cluster_coherence_comparison(
    query = llorens,
    reference = dulken[,dulken$cell_type1!="Ast"]
  )

cluster_coherence_plot <- plot_cluster_coherence_comparison(comparison_results) +
    ggtitle("Reference = Dulken (no astrocytes)\nQuery = Llorens-Bobadilla")
cc_plts <- list()
cluster_coherence_plot -> cc_plts[[1]]
cc_plts[[1]]
```


```{r}
comparison_results <-
  run_cluster_coherence_comparison(
    query = dulken[,dulken$cell_type1!="Ast"],
    reference = llorens
  )
plot_cluster_coherence_comparison(comparison_results) +
  ggtitle("Reference = Llorens-Bobadilla, \nquery = Dulken (no astrocytes)") -> cc_plts[[2]]
cc_plts[[2]]
```


```{r}
comparison_results <-
  run_cluster_coherence_comparison(
    query = llorens,
    reference = dulken
  )
plot_cluster_coherence_comparison(comparison_results) +
  ggtitle("Reference = Llorens-Bobadilla, \nquery = Dulken") -> cc_plts[[3]]

cc_plts[[3]]
```




```{r}
comparison_results <-
  run_cluster_coherence_comparison(
    query = dulken,
    reference = llorens
  )
plot_cluster_coherence_comparison(comparison_results) +
  ggtitle("Reference = Dulken,\nquery = Llorens-Bobadilla") -> cc_plts[[4]]

cc_plts[[4]]
```


```{r, include = F}

cc_plts %>%
  lapply(function(p)
    p + theme(legend.position = "none")
    ) -> cc_plts
cc_plts[["labels"]] <- LETTERS[1:4]
cc_lgnd <- get_legend(cc_plts[[1]] +
                        theme(legend.position = "bottom"))

plot_grid(do.call(what = plot_grid,args = cc_plts),
          cc_lgnd,
          nrow = 2,
          rel_heights = c(10,1)) -> kappa_plt


ggsave(filename = "neurogenesis_kappa.eps",device = "eps",plot = kappa_plt, path = "~/CorGI_figures/neurogenesis",width = 8,height = 8)

```

```{r}
embeddings_mds <- 
  lapply(
    X = gene_sets,
    FUN = function(gene_set) {
      D <- (1 - cor(comb[gene_set, ], method = "spearman")) / 2
      return(cmdscale(D, k = 2))
    }
  )
```


```{r}
emb_name <- "MDS"
embeddings <- embeddings_mds
get_scatterplots(embeddings,
                 batch,
                 cell_type2) -> dr_plots
# 

# dr_plots[[length(dr_plots)+1]] <- legends

dr_plots[["nrow"]] <- 2

dr_plots[["labels"]] <- c("A","B","C","D","E","F")

do.call(plot_grid,dr_plots) -> top_panel
top_panel
```

```{r}
plot_grid(
  plot_grid(
    get_axes_legend(emb_name),
    get_shape_legend(batch, my_shape_palette)
  ),
  get_color_legend(cell_type2, my_color_palette),
  nrow = 2) -> legends
```








```{r,fig.height=2,fig.width=6}
plot_grid(
  legends,
  cluster_coherence_plot + theme(plot.title = element_text(size=10)),
  nrow = 1,
  rel_widths = c(1,2),
  labels = c("","G")) -> bottom_panel

bottom_panel
```




```{r,fig.width=6,fig.height=6}
panel_plot_mds <- plot_grid(top_panel,bottom_panel,nrow=2,rel_heights = c(2,1))
panel_plot_mds
```


```{r, include = F}
ggsave(filename = "neurogenesis_MDS.eps",device = "eps",plot = panel_plot_mds,path = "~/CorGI_figures/neurogenesis",width = 8,height = 8)
```


```{r}
embeddings_mnn_pca <- 
  lapply(
    X = gene_sets,
    FUN = function(gene_set) {
      mnn.out <- mnnCorrect(logcounts(dulken[gene_set,]),logcounts(llorens[gene_set,]))
      t.mnn <<- as.matrix(t(do.call(cbind, mnn.out$corrected)))
      pca.mnn <- prcomp(t.mnn, rank=2)
      return(pca.mnn$x)
    }
  )
```



```{r}
emb_name <- "MNNPC"
embeddings <- embeddings_mnn_pca
get_scatterplots(embeddings,
                 batch,
                 cell_type2) -> dr_plots

dr_plots[["nrow"]] <- 2

dr_plots[["labels"]] <- c("A","B","C","D","E","F")

do.call(plot_grid,dr_plots) -> top_panel
top_panel
```

```{r}
plot_grid(
  get_axes_legend(emb_name),
  get_shape_legend(batch, my_shape_palette),
  get_color_legend(cell_type2, my_color_palette),
  nrow = 3) -> legends
```







```{r,fig.width=8,fig.height=5}
panel_plot_mnn <- plot_grid(top_panel,legends,nrow=1,rel_widths = c(6,1))
panel_plot_mnn
```


```{r, include = F}

ggsave(filename = "neurogenesis_MNN.eps",device = "eps",plot = panel_plot_mnn,path = "~/CorGI_figures/neurogenesis",width = 8,height = 5)
```


